Multi-point Aeropropulsive Design Optimization of a High-Bypass Turbofan Engine

### 2022-11-10 conference-abstract/main.tex
updated figures

\documentclass[]{icas2022}

\usepackage{float}

\usepackage[sort&compress,numbers]{natbib} % For bibtex \citet, \citep
\usepackage{hypernat} % To get natbib to play nicely with hyperref
\usepackage{doi} % For getting hyperlinked DOI in the references

@@

\TitlePaper{Coupled Low-Emission Propulsion Analysis and Optimization of a High Bypass Turbofan Engine}
\AuthorPaper[1]{Peter N. Atma}
\AuthorPaper[2]{Andrew H.~R.~Lamkin}
\AuthorPaper[3]{Joaquim R.~R.~A.~Martins}
\affil[1]{MSE Student, Department of Aerospace Engineering, University of Michigan}
\affil[2]{Ph.D.~Candidate, Department of Aerospace Engineering, University of Michigan}
\affil[3]{Professor, Department of Aerospace Engineering, University of Michigan}

% ==================================================
%	Abstract
% ==================================================
\abstractEnglish{
    Advances in commercial propulsion technology has lead to the development of larger and more advanced high-bypass turbofan engines with larger bypass ratios, pressure ratios, and internal temperatures. 
    With the impending climate crisis, aircraft and engine manufacturers have been pushing for greener fuels and more efficient engines to lower greenhouse and NOx emissions. 
    This has revitalized research into using liquid hydrogen as a fuel for aircraft turbine engines in addition to using known thrust augmentation techniques to reduce NOx emissions and improve fuel efficiency. 
    The high-bypass turbofan study investigates the industry-proposed benefits of using a closed-loop water injection and water vapor recovery system with two types of fuels. 
    We will perform a single point analysis to explore the design space and a multi-point gradient-based optimization to determine and validate the efficiency and emission improvements associated with the proposed engine model. 
    Finally, we will present the results from the design space analysis and multi-point optimization from this model. These developments will accelerate and validate current research into the next generation of fuel-efficient, low emission high-bypass turbofan engines.

}
\keywords{Propulsion, Optimization, Emissions}

@@

% ==================================================
%	Body of Document
% ==================================================
\body  %%% Do not remove

@@

\section{Emission-Propulsion Model Description}
\label{sec:epModel}
% Engine Architecture: Describe the flow path of the engine and establish the mechanical coupling.
We will use a next-generation high-bypass turbofan engine configuration for this work.
Specifically, the NASA advanced technology high bypass geared turbofan engine cycle, referred to as the "N+3" engine, was selected for this research \cite{Hendricks2019}. The N+3 reference cycle represents a notional high bypass ratio geared turbofan that could be available in the 2030–2040 time frame. 
This engine was selected since it has already been modeled in pyCycle as a robust example cycle and includes the advanced engine cycle improvements that were mentioned before such as a large bypass ratio, high pressure ratio, and high combustion temperature. 
This model consists of an inlet that directs ambient air through a fan, followed by a duct that splits the flow into a core flow and a bypass flow, and ends in a bypass nozzle and core nozzle. The fan and low pressure compressor (LPC) are connected to the low pressure turbine (LPT) by the low pressure shaft. 
Similarly, the high pressure compressor (HPC) is connected to the high pressure turbine (HPT) by the high pressure shaft. Along the axial flow path, a the one-dimensional thermodynamic connections are solved using CEA to ensure first the principle governing equations are satisfied. 
We then introduce a water injection component directly before the HPC. 
This will inject pure water vapor into the core stream to reduce the flame combustion temperature. 
The water vapor recovery component is then introduced directly before the core nozzle to recover water vapor from the core stream. 
While in practice this component would require a technical analysis of a condenser in the core stream, this work is only looking at the potential benefit of extracting water vapor and not the recovery method specifically. 
With these components integrated into the N+3 engine, the coupled water inject and water recovery presents a implicit relationship which would need to be iterative solver to match the flow rate entering the engine and exiting the engine. 
The flow and mechanical connections within the engine model with the water injector and water extractor are visualized in Figure \ref{fig:hbtf_cycle}.

@@

% Subsystem Models: Explain the subsystem modeling and coupling in moderate detail

\section{Implementation}
\label{sec:imp}
% Fully coupled model
The propulsion model is constructed in pyCycle which is built on top of OpenMDAO to allow a modular design of the engine components and allows the coupling of other engineering disciplines for analysis \cite{openmdao_2019}. 
For the Jet-A fuel, the NOx correlation will use the P3-T3 method which is commonly a used technique in industry to determine NOx formation in aircraft engines while considering humidity (CITE P3T3). 
This method uses the pressure and temperature of the flow before the combustor to determine NOx formation from a reference design condition of the engine. 
Since not many NOx correlations exist for the NOx formation for the liquid hydrogen fuel, we will use a surrogate model using data computed using Cantera to approximately model the NOx emissions.

@@

\section{Contact Author Email Address}
mailto: patma@umich.edu

\section{Copyright Statement} % update for Aviation instead of ICAS
\begin{small}
	The authors confirm that they, and/or their company or organization, hold copyright on all of the original material included in this paper. The authors also confirm that they have obtained permission, from the copyright holder of any third party material included in this paper, to publish it as part of their paper. The authors confirm that they give permission, or have obtained permission from the copyright holder of this paper, for the publication and distribution of this paper as part of the ICAS proceedings or as individual off-prints from the proceedings.
\end{small}

\bibliographystyle{new-aiaa}
\bibliography{mdolab}

>>>>

\RequirePackage{fix-cm}
\documentclass[conf]{new-aiaa}

\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{siunitx}
\sisetup{group-separator = {,}}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{float}
\usepackage{amsmath}
\usepackage[version=4]{mhchem}
\usepackage{longtable,tabularx}
\usepackage{placeins}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{latexsym}
\usepackage{subcaption}
\usepackage{latexsym}
\usepackage[sort&compress,numbers]{natbib} % For bibtex \citet, \citep
\usepackage{hypernat} % To get natbib to play nicely with hyperref
\usepackage{doi} % For getting hyperlinked DOI in the references

\hypersetup{
    pdfauthor={Peter Atma}, % insert author here
	pdftitle={Aviation 2022 Abstract}, % insert title here
	pdfsubject={}, % insert keywords here
}

@@

\title{Multi-point Aeropropulsive Design Optimization of a High-Bypass Turbofan Engine} % change

\author{Peter Atma\footnote{MS Student, Department of Aerospace Engineering, AIAA Student Member}}
\author{Andrew H.R. Lamkin\footnote{Ph.D.~Candidate, Department of Aerospace Engineering, AIAA Student Member}}
\author{Joaquim R.~R.~A.~Martins\footnote{Professor, Department of Aerospace Engineering, AIAA Fellow}}
\affil{University of Michigan, Ann Arbor, MI, 48109}

% ==================================================
%	Abstract
% ==================================================

@@

\maketitle

\begin{abstract}
	Advances in commercial propulsion technology has lead to the development of larger and more advanced high-bypass turbofan engines with larger bypass ratios, pressure ratios, and internal temperatures.
	With the impending climate crisis, aircraft and engine manufacturers have been pushing for greener fuels and more efficient engines to lower greenhouse and NOx emissions.
	This has revitalized research into using liquid hydrogen as a fuel for aircraft turbine engines in addition to using known thrust augmentation techniques to reduce NOx emissions and improve fuel efficiency.
	The high-bypass turbofan study investigates the industry-proposed benefits of using a closed-loop water injection and water vapor recovery system with two types of fuels.
	We will perform a single point analysis to explore the design space and a multi-point gradient-based optimization to determine and validate the efficiency and emission improvements associated with the proposed engine model.
	Finally, we will present the results from the design space analysis and multi-point optimization from this model. These developments will accelerate and validate current research into the next generation of fuel-efficient, low emission high-bypass turbofan engines.
\end{abstract}

@@

\section{Emission-Propulsion Model Description}
\label{sec:epModel}
% Engine Architecture: Describe the flow path of the engine and establish the mechanical coupling.
We will use a next-generation high-bypass turbofan engine configuration for this work.
Specifically, the NASA advanced technology high bypass geared turbofan engine cycle, referred to as the "N+3" engine, was selected for this research \cite[]{Hendricks2019}. The N+3 reference cycle represents a notional high bypass ratio geared turbofan that could be available in the 2030–2040 time frame.
This engine was selected since it has already been modeled in pyCycle as a robust example cycle and includes the advanced engine cycle improvements that were mentioned before such as a large bypass ratio, high pressure ratio, and high combustion temperature.
This model consists of an inlet that directs ambient air through a fan, followed by a duct that splits the flow into a core flow and a bypass flow, and ends in a bypass nozzle and core nozzle. The fan and low pressure compressor (LPC) are connected to the low pressure turbine (LPT) by the low pressure shaft.
Similarly, the high pressure compressor (HPC) is connected to the high pressure turbine (HPT) by the high pressure shaft. Along the axial flow path, a the one-dimensional thermodynamic connections are solved using CEA to ensure first the principle governing equations are satisfied.
We then introduce a water injection component directly before the HPC.
This will inject pure water vapor into the core stream to reduce the flame combustion temperature.
The water vapor recovery component is then introduced directly before the core nozzle to recover water vapor from the core stream.
While in practice this component would require a technical analysis of a condenser in the core stream, this work is only looking at the potential benefit of extracting water vapor and not the recovery method specifically.
With these components integrated into the N+3 engine, the coupled water inject and water recovery presents a implicit relationship which would need to be iterative solver to match the flow rate entering the engine and exiting the engine.
The flow and mechanical connections within the engine model with the water injector and water extractor are visualized in Figure \ref{fig:hbtf_cycle}.

@@

% Preliminary Results
This model has been tested using Jet-A and liquid hydrogen fuels to determine the improvements in efficiency with injecting water into the core stream.
Preliminary results are shown in Figure \ref{fig:results} and show the relative improvement in thrust specfic energy consumption with water injection for the same engine design with each type of fuel.
We can already see great improvements in the cruise engine efficiency which may be able to be further improved with a full engine optimization.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{JetA-H2_bar_chart_diff.png.pdf}
	\caption{Relative improvement in .}
	\label{fig:results}
\end{figure}

\section{Implementation}
\label{sec:imp}
% Fully coupled model
The propulsion model is constructed in pyCycle which is built on top of OpenMDAO to allow a modular design of the engine components and allows the coupling of other engineering disciplines for analysis \cite{Gray2019a}.
For the Jet-A fuel, the NOx correlation will use the P3-T3 method which is commonly a used technique in industry to determine NOx formation in aircraft engines while considering humidity (CITE P3T3).
This method uses the pressure and temperature of the flow before the combustor to determine NOx formation from a reference design condition of the engine.
Since not many NOx correlations exist for the NOx formation for the liquid hydrogen fuel, we will use a surrogate model using data computed using Cantera to approximately model the NOx emissions.

@@

\bibliography{mdolab,references}
